---
title: "W10 Data Cleaning and Merging"
author: "Meiyao Li"
date: "2023-11-28"
output: html_document
---

# Cleaning, reshaping, merging data
```{r}
library(tidyverse)
```

Each observation is a row (horizontal)
Each variable is a column (vertical)

## 15.2 Date and time
### Duration
For time intervals less than 1hr, we write `MM:SS`, e.g. `58:30`


## 15.3 reshaping data
### 15.3.1 Long form vs. wide form data
Long form: every variable is conceptually different, has more observations
- state, year, drinking %
- has values that do repeat in the first column.
Wide form: variables contain a similar measure
- state, year+drinking%
- has values that do not repeat in the first column.

## Let's get our hands dirty!
This is a *wide form*, it's easier to compare the indicator of diff years or countries, good for humans to understand
```{r}
gdp <- read_delim("../data/oecd-gdp.csv.bz2")
gdp %>%
  sample_n(10)
```

But long form is good for various data processing pipelines, let's turn it into long form!
```{r}
longGdp <- gdp %>% 
  pivot_longer(!c(country, indicator), 
               names_to = "year", 
                  # here year is <character>
               values_to = "gdp") %>% 
  mutate(year = as.numeric(year))
                  # turn year from chr to double <dbl>
longGdp %>% 
  head (7)
```

You can also turn it back to wide form
```{r}
wideGdp <- longGdp %>% 
  pivot_wider(names_from = "year", 
              values_from = "gdp")

wideGdp %>% 
  head (7)

```


```{r}
p <- data.frame(pregnant = c("Yes", "No"),
                male = c(NA, 25),
                female = c(11,18))

p
```

```{r}
pl <- p %>% 
   pivot_longer(c(male, female),
           	names_to = "sex",
           	values_to = "count")

pl
```
```{r}
pw <- pl %>% 
   pivot_wider(names_from = sex,
           	values_from = count)
pw
```